Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-09-26 Rmarkdown/datasets.Rmd
Closes #22
Closes #21

1. Genetic variants were called using the `bissnp_easy_usage.pl` script included with `Bis-SNP (v0.82.2)`.
2. M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set. These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e. $|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (_read-1_ and _read-2_ analysed separately where applicable).
3. CpG methylation calling was performed using `methtuple (v1.4.0)` for m-tuples $\text{m} = 1, \ldots, 8$. CpG 2-tuples were called both with and without the `--all-combinations` flag; all other tuple sizes were called without the `--all-combinations` flag.
4. Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see Section \ref{sec:MethylationTuples}). Specifically, a `MethPat` object was created for each of the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets for 1-tuples, 2-tuples and 2ac-tuples (2-tuples with the  `--all-combinations` flag in `methtuple` set), 3-tuples and 4-tuples. I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.

@@

There are a four 'mini datasets' within the Lister data that I make some use of in my thesis. The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper. The _ADS_ sample, a human adipose tissue cell line, is the 'founder' of this mini dataset. The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line. The _ADS-adipose_ sample are "adipocytes derived from the _ADS_ cells through adipogenic differentiation conditions". The _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.

>>>>

1. Genetic variants were called using the `bissnp_easy_usage.pl` script included with `Bis-SNP (v0.82.2)`.
2. M-bias was estimated using `bismark_methylation_extractor` with the `--mbias_only` flag set. These output files were then analysed using the  `MethylationTuples` R package (see section \ref{sec:MethylationTuples}) and all read-positions with an CpG normalised read-position methylation level ($nrpml$) more than $0.03$ from the median, i.e. $|nrpml_{CpG} - median(nrpml_{CpG})| > 0.03$, were excluded from methylation calling (_read-1_ and _read-2_ analysed separately where applicable).
3. CpG methylation calling was performed using `methtuple (v1.4.0)` for m-tuples $\text{m} = 1, \ldots, 8$. CpG 2-tuples were called both with and without the `--all-combinations` flag; all other tuple sizes were called without the `--all-combinations` flag. The following `methtuple` flags were also used: `--methylation-type CG --ignore-duplicates --min-mapq 0 --overlap-filter XM_ol --ignore-duplicates`.
4. Sample-level m-tuples were combined at the dataset-level using the `MethylationTuples` R package (see Section \ref{sec:MethylationTuples}). Specifically, a `MethPat` object was created for each of the _EPISCOPE_, _Lister_, _Seisenberger_ and _Ziller_ datasets for 1-tuples, 2-tuples and 2ac-tuples (2-tuples with the  `--all-combinations` flag in `methtuple` set), 3-tuples and 4-tuples. I did not create `MethPat` objects for m-tuples with $\text{m} > 4$ because the data are too sparse at this larger sizes to be generally useful.

@@

There are four 'mini datasets' within the Lister data that I make some use of in my thesis. The first I refer to as the _Lister-ADS_ data and includes samples _ADS_, _ADS-adipose_ and _ADS-iPSC_, all from the 2011 paper. The _ADS_ sample, a human adipose tissue cell line, is the 'founder' of this mini dataset. The _ADS-adipose_ and _ADS-iPSC_ are both derived from the _ADS_ cell line. The _ADS-adipose_ sample are "adipocytes derived from the _ADS_ cells through adipogenic differentiation conditions". The _ADS-iPSC_ cell line is an induced pluripotent stem cell line derived from _ADS_.

### 2015-09-26 Rmarkdown/methsim.Rmd
Closes #21

Almost all downstream analyses are reference-based (see Section \ref{sec:methylated_or_unmethylated}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.

>>>>

Almost all downstream analyses are reference-based (see Section \ref{sec:methylation_calling}), so it is desirable to know how these methods perform with respect to the relevant reference genome, not a simulated genome in which the location of methylation loci vary from simulation to simulation.

### 2015-09-26 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Closes #21

\subsection{Methylated or unmethylated?}\label{sec:methylated_or_unmethylated}

@@

@@

@@

- Whole-genome sequencing or genotyping of the sample
- Calling genetic variations directly from the bisulfite-sequencing data
- Excluding sites of known genetic variation

\subsubsection{Whole-genome sequencing or genotyping of the sample}\label{sec:wgs_or_genotyping}

@@

\subsubsection{Calling genetic variations directly from the bisulfite-sequencing data}\label{sec:wgbs_variant_calling}

@@

\subsubsection{Excluding sites of known genetic variation}\label{sec:known_genetic_variation}

@@

\subsection{Determining the context or methylation type}\label{sec:methylation_context}

@@

`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`. The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program. There are many useful options to filter reads and read-positions. Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting the bases in the overlapping region. `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads. This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}. Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.

@@

I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for more than $40$ whole-genome bisulfite-sequencing samples. Figure \ref{fig:methtuple_runtime} shows the distribution of running times and Figure \ref{fig:methtuple_memory} the maximum memory usage across all the samples from the EPISCOPE, Lister and Ziller datasets. For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set. Note that the total number of samples is $49$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption[\texttt{methtuple} memory usage]{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{{-}{-}all-combinations} was set. Note that the total number of samples is $49$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_memory}
\end{figure}

@@

>>>>

@@

\subsection{Considerations}

There are several issues that must be carefully considered when performing methylation calling, including filtering of reads and biases, choosing and refining the reference sequence, and determining the context or methylation type of the cytosine.

\subsubsection{Filtering of reads and bases}

@@

Although incomplete bisulfite-conversion is a well-recognised issue, most analysis pipelines don't attempt to account for this during methylation calling[^incomplete-conversion]. An exception is `Bis-SNP` \citep{Liu:2012ge} which has an algorithm to exclude bases suspected of suffering from incomplete bisulfite-conversion at the 5' end of Illumina-generated reads.

[^incomplete-conversion]: Instead, incomplete bisulfite-conversion is usually incorporated into calculations to estimate the methylation level at a given cytosine (see Section \ref{sec:beta-value_interpretation}). The bisulfite-conversion rate may be estimated by analysing cytosines that are expected to be unmethylated, such as those in the chloroplast genome \citep{Lister:2008bh} or a spike-in control of lambda phage DNA \citep{Lister:2009hy}.

\subsubsection{Choosing and refining the reference sequence}

@@

1. Whole-genome sequencing or genotyping of the sample
2. Calling genetic variations directly from the bisulfite-sequencing data
3. Excluding sites of known genetic variation

@@

@@

@@

\subsubsection{Determining the context or methylation type}

@@

`methtuple` performs methylation calling for a single `BAM` file generated by `Bismark`. The user is required to specify the size of the tuples (`--m`), and the methylation type (`--methylationType`) for each run of the program. There are many useful options to filter reads and read-positions. Apart from the standard quality filters, `methtuple` is careful when processing paired-end reads to only count the base from one of the reads in any overlapping paired-end reads to avoid double-counting the bases in the overlapping region[^overlapping-mates]. `methtuple` also allows the user to filter out specific read-positions rather than wholesale filtering of the ends of reads. This is particularly useful for samples where there is a 'spike' in the M-bias plot, such as that shown in Figure \ref{fig:E18VA_M-bias}. Such a spike can be filtered out without also being forced to also filter out additional upstream read-positions that are not affected by M-bias.

[^overlapping-mates]: `methtuple` has several options for handling overlapping mates of paired-end reads via the `--overlap-filter` flag. The default method is to ignore any read-positions in the overlapping region where the methylation calls disagree.

@@

I have used `methtuple` to perform methylation calling at CpG m-tuples, $m = 1, \ldots, 8$, for more than $40$ whole-genome bisulfite-sequencing samples. Figure \ref{fig:methtuple_runtime} shows the distribution of running times, Figure \ref{fig:methtuple_memory} the maximum memory usage across, and Figure \ref{fig:methtuple_file_sizes} the output file sizes, for all the samples from the EPISCOPE, Lister, and Ziller datasets. For each sample, each chromosome was processed using a single core on one of the shared-use servers in the Bioinformatics Division (see Table \ref{tab:unices} in the Appendix for details of these machines).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_runtime.pdf}
\caption[\texttt{methtuple} running time]{The running times are the `User time' reported by \texttt{GNU time} converted from seconds to minutes. The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set. Note that the total number of samples is $48$ because each of the Ziller sequencing runs is separately counted (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_runtime}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_memory.pdf}
\caption[\texttt{methtuple} memory usage]{The maximum memory usage is the `Maximum resident set size' reported by \texttt{GNU time} converted from kilobytes to gigabytes. These values are divided by four to fix bug in how \texttt{GNU time} reports the maximum memory usage (\url{https://bugzilla.redhat.com/show_bug.cgi?id=703865}). The suffix `ac' means that the option \texttt{{-}{-}all-combinations} was set. Note that the total number of samples is $48$ because each of the Ziller sequencing runs is separetely counted (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_memory}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/methtuple_file_sizes.pdf}
\caption[\texttt{methtuple} file sizes]{The sizes of the output file when compressed with gzip. The suffix `ac' on the tuple size means that the option \texttt{{-}{-}all-combinations} was set. Note that the total number of samples is $48$ because each of the Ziller sequencing runs is separately counted (see Chapter \ref{chap:datasets} for details).}
\label{fig:methtuple_file_sizes}
\end{figure}

@@

The regular structure of the output file means that these are particularly compressible. The size of the output file is almost always less than 1 GB when compressed with `gzip`.
